---
title: "Stat 151A Lab 10"
author: "Bryan Liu"
date: "October 26, 2018"
output:
  html_document:
    toc: yes
    toc_float:
      collapsed: no
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  cache=TRUE, warning=FALSE, error=FALSE, message=FALSE, echo=TRUE, 
  results = 'markup', dev='png', dpi=150
)
```

```{r}
library(tidyverse)
library(MASS)
library(leaps)

set.seed(245234)
```

# Statistics is made up and nothing is real

"If you torture the data long enough, it will confess." -- Ronald Coase

This example is adapted from Freedman 1983[^1]. 

[^1]: Freedman, D. A. (1983) "A note on screening regression equations." The American Statistician, 37, 152–155

Suppose we have 500 observations with each $x_i \in \mathbb{R}^{100}$, and $y_i \in \mathbb{R}$. We draw $x_i \sim \mathcal{N}(0, I_{100 \times 100})$ and $y_i \sim \mathcal{N}(0, 1)$. In other words, OUR DATAFRAME IS JUST RANDOM, INDEPENDENT NOISE. Literally every entry in our dataframe is independent, and most crucially, the $y$'s have no relationship to the $x$'s. 

We will run a regression on this dataset, and see what we get. 

```{r}
n <- 500 # number of observations
p <- 100 # dimension of features

# draw X 
X <- matrix(rnorm(n * p), nrow = n)

# draw y
y <- rnorm(n)

# create dataframe
data_noise <- cbind(y, X)
data_noise <- as.data.frame(data_noise)

# fit regression
full_model <- lm(y ~ ., data_noise)
full_model_summary <- summary(full_model)

full_model_summary
```
What are the true values for $\beta_0, ..., \beta_{100}$ in our data generating process? 

What doe the $F$ statistic say? Is this consistent with our data generating process?

Lets see how many $\hat\beta_i$'s were significantly different from 0. 
```{r}
# get p-values
p_vals <- coef(full_model_summary)[, 'Pr(>|t|)']

# print how many were less than 0.05 
print(sum(p_vals < 0.05))
```

Is this surprising?

Ok, so now lets do model selection by including only the coefficients with the top 3 p-values in the model. 
```{r}
# pick out top three coefficients
which_top3 <- order(p_vals)[1:3]
print(p_vals[which_top3])

# and only run the regression on the coefficients with the top 3 p-values
data_noise_selected <- data_noise[,  c('y', names(p_vals[which_top3]))]
summary(lm(y ~ ., data = data_noise_selected))

```
What does the F-statistic say now??? (Remember, under our data generating process, was the null hypothesis true or false?)

Fine, lets try something fancier than just picking coefficients with large p-values. In class we talked about doing stepwise search. Lets do a backwards search, with AIC as our criterion. R has a nice function for this. It'll suffice for our demonstration. 
```{r}
# this line takes a minute
aic_selected_model <- stepAIC(full_model, direction = 'backward', 
                              trace = FALSE)
# print the result
summary(aic_selected_model)
```

Observations about $F$ and/or $T$ statistics?

# Anyyyyyway ...
Lets pretend we did not see that and continue with this lab. 

Note that that was a caution about *inference*. We can always use linear regression and whatever selection techniques we choose, if we are only aiming for *prediction*. 

## (Information) Criteria
In our model, we have a total of $p$ explainatory variables. Suppose we decide only to include a subset of $p' \leq p$ variables in our model. **How many possible models with $p'$ variables are there?**

Suppose we can search through all the models with $p'$ explainatory variables. We need some criterion to determine which is best. 

A quantity that shows up in many of the information criteria discussed in class is the log-likelihood of the linear model, with normal errors, at the estimates $\hat\beta$ and $\hat\sigma^2$. 

Suppose $y \sim \mathcal{N}(X\beta, \sigma^2 I_{n\times n})$. Then the log-likelihood is given by 
\begin{align}
\log p(y ; \beta, \sigma^2)
= - \frac{n}{2} \log(2 \pi \sigma^2) - \frac{\|y - X\beta\|^2}{2 \sigma^2}
\end{align}

**Check that $\hat\beta_{MLE} = (X^TX)^{-1}X^Ty$, and $\hat\sigma^2_{MLE} = RSS / n$ is the MLE**, that is, these formulae for $\hat\beta_{MLE}$ and $\hat\sigma_{MLE}^2$ maximize the log-likelihood. (Note that $\hat\beta_{MLE}$ is our usual formula obtained by solving the least squares formula; however, $\hat\sigma^2_{MLE} = RSS / n$ is not the familiar unbiased estimate for the variance. )

So now let us list the criteria that we might use to select which model is best. Let $m$ denote the model, with $p(m)$ being the number of explainatory variables in the model: 

* Adjusted $R^2$: 

\begin{align}
(\text{Adj. $R^2$})(m) := 1 - \frac{\text{RSS}(m) / (n - p(m) - 1)}{\text{TSS} / (n - 1)}
\end{align}

* AIC: 

\begin{align}
\text{AIC}(m)
&:= - 2 \log(\text{max value of likelihood in $m$}) + 2(\text{number of parameters in $m$})
\\
&= n \log(2 \pi e) + n \log \left(\frac{\text{RSS(m)}}{n}\right) + 2(p(m)+1)
\end{align}

We compare AIC$(m)$ for all models $m$, so we can drop the constant $n\log(2\pi e)$ since it does not depend on $m$. Some implementations have different terms in computing their AIC . But they should differ by only an additive constant. 

* BIC: 

\begin{align}
\text{BIC}(m)
&:= - 2 \log(\text{max value of likelihood in $m$}) + \log(n) \cdot (\text{number of parameters in $m$})
\\
&= n \log(2 \pi e) + n \log \left(\frac{\text{RSS(m)}}{n}\right) + \log(n) \cdot (p(m)+1).
\end{align}

We can again drop the constant term. How does BIC penalize complexity $p(m)$ compared to AIC?

* Mallows $C_p$: Suppose that the data did come from the model $y = X\beta + \epsilon$, $\epsilon \sim \mathcal{N}(0, I_{n\times n})$. The idea behind the Mallows $C_p$ score is that the $C_p$ score is a proxy of the quantity 

\begin{align}
E\|H(m) Y - X\beta\|^2
\end{align}

We would like to choose $m$ to minimize this expectation. In class, we computed this to be 
\begin{align}
E\|H(m) Y - X\beta\|^2= \sigma^2(p(m) + 1) + \beta^T X^T(I - H(m))X\beta
\end{align}

**Show that this equals $\sigma^2(p(M) + 1)$ when $M$ is the full model**. 


We claim that 

\begin{align}
RSS(m) - \sigma^2(n - 2 - 2p(m))
\end{align}
is an unbiased estimate of the expectation. **Show.** 

We divide through by $\sigma^2$, and replace $\sigma^2$ with $\hat\sigma^2$. This results in Mallow's $C_p$ score: 

\begin{align}
C_p(m) := \frac{RSS(m)}{\hat\sigma^2} - (n - 2 - 2p(m))
\end{align}


So now, to do model selection, we pick our favorite criteria, and optimize the criterion in two stages: 

1. For each subset size $p' = 0, ..., p$, find the model with exactly $p'$ explainatory variables with the smallest RSS (for these criteria, why does it suffice to just minimize RSS in step 1?). 

2. For these $p + 1$ models, pick the one with the smallest criterion (AIC/BIC/etc) value. 

## Supernova dataset

We consider a dataset from Brad Efron and Trevor Hastie's book, *Computer Age Statistical Inference*. The data measures the brightness of 39 supernova, as a function of $p = 10$ spectral energies. Note that the brightnesses (the $y$) have been centered to have zero mean, and the $x$'s are scaled to have zero mean and unit variance. 

```{r}
# load data 
supernova <- read.table("http://web.stanford.edu/~hastie/CASI_files/DATA/supernova.txt", header=TRUE)

head(supernova)
```

The goal is to use a regression model to predict supernova brightness from the spectral energies. Lets hold out 10 of the 39 observations as a test set. 
```{r}
# Do train/test split
set.seed(4534534)
sample_10 <- sample(dim(supernova)[1], 10)

supernova_train <- supernova[-sample_10, ]
supernova_test <- supernova[sample_10, ]
```

```{r}
# fit on training set 
lm_fit <- lm(Magnitude ~ . -1, data = supernova_train)

# examine predictive accuracy on the training set 
predict_train_data <- predict(lm_fit, data = supernova_train)
R2_train <- cor(predict_train_data, supernova_train[, 'Magnitude'])**2
cat('R^2 on training data: ', R2_train)
```
Not bad, right?
```{r}
# plot observed vs predicted values in the training set 
ggplot() + 
  geom_point(aes(x = predict_train_data, y = supernova_train[, 'Magnitude']), 
             color = 'blue') + 
  ylab('observed magnitude') + xlab('predicted magnitude') + 
  geom_abline(slope = 1, color = 'red')
```

```{r}
pred_test_data <- predict(lm_fit, newdata = supernova_test)
R2_test <- cor(pred_test_data, supernova_test[, 'Magnitude'])**2
# get R^2 of test data  
cat('R^2 on test data: ', R2_test)

```

What happened?

Let us do variable selection and search for which of the 10 spectal energies are actually important for prediction. We will use the ``leaps`` package in R, and use Mallow's Cp score. 

```{r}
leaps_output <- leaps(y = supernova$Magnitude, x = supernova[, 1:10], 
                      method = 'Cp', nbest = 1, int = FALSE)

print(leaps_output)
```

Using the selected explainatory variables, lets fit again using the training data. 
```{r}
# Probably an easier way to do this. 
# was too lazy to dig into leaps documentation
# so I'll do it myself
# This function returns a dataframe with only the X selected columns (and the y)
get_selected_x_from_leaps <- function(leaps_ouput, supernova){
  # get the columns that minimizes Cp
  which_model <- which.min(leaps_output$Cp)
  which_X_columns <- leaps_output$which[which_model, ]
  
  # create new dataset with only selected columns
  supernova_selected_x <- cbind(supernova[, c(which_X_columns)], supernova[, 11])
  names(supernova_selected_x)[length(names(supernova_selected_x))] <- 'Magnitude'
  
  return(supernova_selected_x)
}

# get selected fit
selected_supernova_train <- get_selected_x_from_leaps(leaps_ouput, supernova_train)
lm_fit_selected <- lm(Magnitude ~ . -1, data = selected_supernova_train)

# check R^2 on train data
pred_train_data <- predict(lm_fit_selected, newdata = selected_supernova_train)
R2_train <- cor(pred_train_data, selected_supernova_train[, 'Magnitude'])**2
cat('R^2 on training data: ', R2_train)
```
The fit on the test data with selected columns?
```{r}
# check R^2 on test data 
selected_supernova_test <- get_selected_x_from_leaps(leaps_ouput, supernova_test)
pred_test_data <- predict(lm_fit_selected, newdata = selected_supernova_test)
R2_test <- cor(pred_test_data, selected_supernova_test[, 'Magnitude'])**2
cat('R^2 on test data: ', R2_test)

```